Parity games in pushdown systems with parameters

### 2023-06-30 reachability/article.tex
up

In particular, two player games can be used to
 represent the
interaction of a program with some environment, with the first player representing the program while the second player represents the environment. A winning condition then expresses a property required to hold however the environment behaves. Finding a winning strategy for the first player
thus provides some way 
 to ensure that the desired 
property, as expressed by the winning condition,
always holds \cite{arnold2003games}.
Games on the graphs of pushdown systems have been extensively studied.
It is known from \cite{walukiewicz1996pushdown} that deciding the winner of a parity game on a graph of a pushdown system is an ${\sf EXPTIME}$-complete problem.

{\bf Parametric constraints.}
 Parametric pushdown systems provide a  formalism to
reason about 
recursive programs
making use of parametric constraints. 
Recently, different variants of parametric pushdown systems have been introduced in the literature 
\cite{hague2011parameterised, esparza2016parameterized, fortin2017model} 
however as
parameterized asynchronous shared-memory systems.
These consist of a leader pushdown system
and arbitrarily many identical contributor pushdown systems, communicating via a shared memory
in the form of a register which can take finitely many values.
They have applications
in the dataflow analysis of concurrent programs~\cite{kahlon2008parameterization}. 
We consider in this article a different approach to extending pushdown systems with parameters, by allowing equality test of the stack against parameters.
Perhaps a similar model consists in 
pushdown systems 
with transitions that are conditioned by regular conditions on their stack content, which
can in particular
test the
stack content against
specific words. 
They can be used to ensure that some word over the stack alphabet appears in the configurations of a run, but only for a specific value (or, rather, specific regular languages).
They have 
been used to  obtain 
model checking algorithms for LTL and CTL$^*$ model checking
for
pushdown systems~\cite{finkel1997direct}.
Instead
of checking the stack content against regular conditions, or
against specified word values,
we consider
checks against unspecified word values. These will be instantiated using parameters,
and compared against using parametric guards.
This allows for example to ensure equality between two stack contents appearing in two distincts configurations in a run.

{\bf Related models.}
Several generalizations of the pushdown systems 
have been studied in depths, including
 ordered multi-pushdown systems \cite{breveglieri1996multi, atig2012model}, annotated higher-order pushdown systems \cite{maslov1976multilevel, broadbent2012saturation}, and
strongly normed multi-pushdown systems \cite{czerwinski2012reachability}.
Extensions of pushdown systems with storage for later comparisons
have been introduced for instance as register pushdown systems. 
Such automata possess registers in addition to their stack, and can keep data values in both.
Register pushdown systems have been shown to have applications for malware detection and XML schema checking \cite{senda2021forward, senda2021ltl}. 
In theory, since registers can be unfilled and later be filled again an unbounded number of times, the number of different data values a pushdown register systems can store in its registers in a run is unbounded.  
In \cite{murawski2017reachability} it was shown however that a register pushdown automaton can
only really ``remember'' at most  $3r$ data values, where $r$ is the number of registers,
i.e. for any run of a register pushdown automaton with $r$ registers there exists an equivalent run 
 with the same initial and final conﬁgurations, but in which every configuration contains register assignments drawn from only $3r$ elements. This  
 leads to
 the
  question
  of
   how useful the ability to unfill and later refill registers really is compared to a model that would only specify valuations once.

@@

{\bf Overview. }
In Section 2, we provide general notations and preliminary definitions. 
Section 3 will deal with the introduction of  parametric pushdown games.
% Section 4 is devoted to using pebble alternating tree systems with parity acceptance condition to solve parametric pushdown system parity games. We leave the proof of decidability itself  for Section 5.
Section 4 is Reduction to higher-order pushdown systems
Section 5 at last shows a nonelementary lower bound for the problem.

>>>>

In particular, two player games can be used to
 represent the
interaction of a program with some environment, with the first player representing the program while the second player represents the environment. A winning condition then expresses a property required to hold however the environment behaves. Finding a winning strategy for the first player
thus provides some way 
 to ensure that the desired 
property
always holds \cite{arnold2003games}.
Games on the graphs of pushdown systems have been extensively studied.
It is known from \cite{walukiewicz1996pushdown} that deciding the winner of a parity game on a graph of a pushdown system is an ${\sf EXPTIME}$-complete problem.

{\bf Parametric constraints.}
% Parametric pushdown systems provide a  formalism to reason about  recursive programs making use of parametric constraints. 
For certain types of computer programs, such as embedded systems, under-specified constraints are necessary rather than only considering concrete specifications.
Unspecified values represented by parameters have been used
for instance in parametric timed automata~\cite{AHV93-stoc} and parametric one-counter automata~\cite{haase2012complexity, lechner2015complexity}.
Verification of the desired properties of the system is then performed without concrete values.
 We consider in this article a similar approach, only applied to pushdown systems. Rather than comparing a counter against a parametric number, we allow comparison of a stack content against a parametric word.
  Perhaps a similar model consists in  pushdown systems with transitions that are conditioned by regular conditions on their stack content, which can in particular test the stack content against specific words. 
They can be used to ensure that some word over the stack alphabet appears in the configurations of a run, but only for a specific value (or, rather, specific regular languages).
They have  been used to  obtain  model checking algorithms for LTL and CTL$^*$ model checking
for pushdown systems~\cite{finkel1997direct}.
Instead of checking the stack content against regular conditions, or against specified word values, we consider checks against unspecified word values. These will be instantiated using parameters, and compared against using parametric guards.
This allows for example to ensure equality between two stack contents appearing in two distincts configurations in a run.

{\bf Related models.}
Recently, different variants of parametric pushdown systems have been introduced in the literature 
\cite{hague2011parameterised, esparza2016parameterized, fortin2017model} 
however as
parameterized asynchronous shared-memory systems.
These consist of a leader pushdown system
and arbitrarily many identical contributor pushdown systems, communicating via a shared memory
in the form of a register which can take finitely many values.
They have applications
in the dataflow analysis of concurrent programs~\cite{kahlon2008parameterization}. 
We consider in this article a different approach to extending pushdown systems with parameters, by making them appear directly in equality tests against the stack.

% Several generalizations of the pushdown systems  have been studied in depths, including ordered multi-pushdown systems \cite{breveglieri1996multi, atig2012model}, annotated higher-order pushdown systems \cite{maslov1976multilevel, broadbent2012saturation}, and strongly normed multi-pushdown systems \cite{czerwinski2012reachability}.
Extensions of pushdown systems with storage for later comparisons
have been introduced for instance as register pushdown systems. 
Such automata possess registers in addition to their stack, and can keep data values in both.
Register pushdown systems have been shown to have applications for malware detection and XML schema checking \cite{senda2021forward, senda2021ltl}. 
In theory, since registers can be unfilled and later be filled again an unbounded number of times, the number of different data values a pushdown register systems can store in its registers in a run is unbounded.  
In \cite{murawski2017reachability} it was shown however that a register pushdown automaton can
only really ``remember'' at most  $3r$ data values, where $r$ is the number of registers,
i.e. for any run of a register pushdown automaton with $r$ registers there exists an equivalent run 
 with the same initial and final conﬁgurations, but in which every configuration contains register assignments drawn from only $3r$ elements. This  
 leads to
 the
  question
  of
   how useful the ability to unfill and later refill registers really is compared to a model that would only specify valuations once.

@@

% {\bf Overview. }
% In Section 2, we provide general notations and preliminary definitions. 
% Section 3 will deal with the introduction of  parametric pushdown games.
% Section 4 is Reduction to higher-order pushdown systems
% Section 5 at last shows a nonelementary lower bound for the problem.
